---
title: "Reproducible Research Project 1"
author: "Papajohnco"
date: "12/12/2021-12/28/2021"
output: html_document
---

## Load dplyr, ggplot2, and knitr packages then call the libraries to this session
```{r}
library(knitr)
library(dplyr)
library(ggplot2)
library(lattice)
```

## Global enable to alway show code chunks in output

```{r}
opts_chunk$set(echo = TRUE)
```

## Set working directory
```{r}
setwd("C:/Users/jcomaiii/Desktop/Coursera Reproducible Research/Reproducible-Research-Repo")
```

## Read the data into Rstudio using the read.csv function and assign a object name "data_rows"

```{r}
data_rows <- read.csv("activity.csv")
```
## Remove NA from data and assign a object name "data" *could also use na.rm = TRUE

```{r}
data <- data_rows [ with (data_rows, { !(is.na(steps)) } ), ]
```

## View first 30 rows without "NA" to ensure that NA is removed

```{r}
head(data, 30)
```

## Adding total steps each day by date using aggregate function

```{r}
aggregate(x = data$steps, by = list(data$date), FUN = sum)
```

## could also calculate total number of steps taken per day as follows

```{r}
by_day <- group_by(data, date)
steps_by_day <- summarise(by_day, total = sum(steps))
steps_by_day
```

## Assign total number of steps taken each day to variable x

```{r}
x <- aggregate(x = data$steps, by = list(data$date), FUN = sum)
```

## Histogram of the total number of steps taken each day

```{r}
hist(steps_by_day$total, main="Histogram of total number of steps taken each day", 
     xlab="Total number of steps taken each day")
```

## Calculate and report the mean and median of the total number of steps taken per day using summary feature

```{r}
summary(steps_by_day)
```
## Preprocessing data for time series plot (i.e. type = “l”) 

```{r}
steps_by_interval <- aggregate(steps ~ interval, data, mean)
```

## Create the time series plot

```{r}
plot(steps_by_interval$interval, steps_by_interval$steps, type='l', 
     main="Average number of steps over all days", 
     xlab="Interval", 
     ylab="Average number of steps")
```
## Process the data to determine which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_steps_row <- which.max(steps_by_interval$steps)
```

## Calculate the interval that has the average max steps row value

```{r}
steps_by_interval[max_steps_row, ]
```
## Therefore the interval that contains the max number of steps is 835, which has average max of just over 206 steps

## Imputing missing values. ie. Calculate and report the total number of missing values in the dataset 

```{r}
sum(is.na(data_rows))
```
## The total number of rows with NA is 2304

## Strategy to add in all the missing rows into the dataset using mean/median for the day. 

```{r}
missingIndex<-is.na(data_rows[,1])
```


```{r}
m<-mean(steps_by_interval$steps)
```

## Create a new dataset with the missing (imputed) data filled in.

```{r}
activity1<-data_rows
activity1[missingIndex,1]<-m
head(activity1)
```

## Histogram of the total number of steps taken each day after imputed data is included

```{r}
totalStepsByDay1<-aggregate(steps~date, activity1, sum)
hist(totalStepsByDay1$steps, xlab="Total Number of Steps per day", ylab="Number of Days", main="Number of Steps taken each day after imputed data is included")
```


```{r}
total_Steps_By_Day1<-aggregate(steps~date, activity1, sum)
```

## Mean after imputed steps added

```{r}
mean_after_imputed_steps<-mean(total_Steps_By_Day1$steps)
mean_after_imputed_steps
```

## Median after imputed steps added

```{r}
median_after_imputed_steps<-median(total_Steps_By_Day1$steps)
median_after_imputed_steps
```
## Using the new dataset with imputed data "activity1", create new variables "weekday" and "weekend"
## If day is Saturday or Sunday then it's weekend, else it's a weekday
## Create these variables in "activity2"

```{r}
activity1$date<-as.Date(activity1$date)
activity2<-activity1%>%
  mutate(daytype= ifelse(weekdays(activity1$date)=="Saturday"|weekdays(activity1$date)=="Sunday", "Weekend", "Weekday"))
head(activity2, 10)
```

## Now create a panel plot comparing the average number of steps taken per 5 minute interval across weekends & weekdays

```{r}
averageStepByDayTypeAndInterval<-activity2%>%
  group_by(daytype, interval)%>%
  summarize(averageStepByDay=sum(steps))
```


```{r}
with(averageStepByDayTypeAndInterval, 
      xyplot(averageStepByDay ~ interval | daytype, 
      type = "l",      
      main = "Average Number of Steps across Weekdays & Weekends",
      xlab = "Daily Intervals",
      ylab = "Average Number of Steps"))
```
